---
title: "Causal Effect with interference"
author: "Sohyeon Kim"
output: 
  github_document:
    pandoc_args: --webtex
---


```{r warning=FALSE, message = FALSE}
library(tidyverse)
library(pander)
```


# 1. Simulation

This is a scenario for binary outcomes, heterogenoeus group between groups. 

## 1-1. Step 1 

Of the 4000 individuals in the population, $x_{ij}$ is generated as follows :

$$x_{ij} = \begin{cases}0 &\text{for randomly selected 480 samples}\\1 &\text{for another 480 random samples}\\2&\text{for remaining individuals} \end{cases}$$


```{r}
set.seed(0)
x0 <- rep(0, 480)
x1 <- rep(1, 480)
x2 <- rep(2, 4000-480*2)

x <- sample(c(x0, x1, x2))
```



Potential outcome is defined as follows : 

$$y_{ij}(z_{ij}, \alpha_{ij})=\begin{cases} g_iz_{ij}&\text{if }x_{ij}=2\text{ and }i=1,2\\g_i(1-z_{ij})&\text{if }x_{ij}=2\text{ and }i=3,4\\x_{ij}&\text{otherwise}\end{cases}$$

## 1-2. Step 2

Groups were assigned $\alpha_0, \alpha_1$ and individuals assigned z = 1,0 using two stage permutation randomization. 

```{r}
g <- sample(c(1, 1, 0, 0))

z <- list()
for(i in 1:length(g)) {
  if(g[i] == 1) {
    z[[i]] <- sample(c(rep(0, 500), rep(1, 500)))
  } else {z[[i]] <- sample(c(rep(0, 800), rep(1, 200)))}
}


data <-  bind_cols(z) %>%
  `colnames<-`(1:4) %>%
  gather(key = "group_idx", value = z) %>%
  bind_cols(data.frame(g = rep(g, each = 1000), x = x))

y <- rep(NA, 4000)
for(i in 1:nrow(data)) {
  if(data$x[i] == 2) {
    y[i] <- ifelse(data$group_idx[i] %in% 1:2, data$g[i]*data$z[i], data$g[i]*(1-data$z[i]))
  } else {y[i] <- x[i]}
}

data <- mutate(data, y = y)
```

## 1-3. Step 3

Calculate various causal effect estimators defined in the paper. 

```{r}
summary <- data %>%
  select(-x) %>%
  group_by(group_idx) %>%
  split(.$group_idx) %>%
  lapply(FUN = function(x) data.frame(Total_treated = sum(x$z), Cases_treated = sum(x$z*(x$y)), 
                                      Total_control = sum(1 - x$z), Cases_control = sum((1-x$z)*x$y))) %>%
  bind_rows(.id = "group_idx") %>%
  mutate(g_i = g) %>%
  select(group_idx, g_i, Total_treated, Cases_treated, Total_control, Cases_control)

K <- summary$Total_treated
N <- length(g)
C <- sum(g)

summary %>% pander()
```



$\text{Total treated }= \sum_jZ_{ij} \text{ when }Z_{ij} = 1$

$\text{Cases treated }= \sum_jZ_{ij}Y_{ij}(Z_{i}) \text{ when }Z_{ij} = 1$

$\text{Total treated }= \sum_j(1-Z_{ij}) \text{ when }Z_{ij} = 0$

$\text{Cases treated }= \sum_j(1-Z_{ij})Y_{ij}(Z_{i}) \text{ when }Z_{ij} = 0$

$g_i=1$ when group assignment of group i is $\psi$

$g_i=0$ when group assignment of group i is $\phi$

# 2. Point Estimate

### Direct Effect

```{r}
data_CE_D <- summary %>% mutate(PO_1 = Cases_treated/Total_treated, 
                   PO_0 = Cases_control/Total_control) %>%
  select(group_idx, PO_1, PO_0) %>% 
  split(.$group_idx) %>%
  lapply(FUN = function(x) data.frame(CE_D = x$PO_0 - x$PO_1)) %>%
  bind_rows(.id = "grou_idx") %>%
  mutate(g = g)

CE_D <- data_CE_D %>%
  group_by(g) %>%
  summarise(CE_D = mean(CE_D)) 

pander(CE_D)
```


$$\bar{CE}^D(\psi)= \sum_{i=1}^2\frac{\bar{CE}_i^D(\psi)}{2} = -0.749$$

$$\bar{CE}^D(\phi)= \sum_{i=3}^4\frac{\bar{CE}_i^D(\phi)}{2} = 0.00625$$



### Indirect Effect

```{r}
summary %>% mutate(PO_0 = Cases_control/Total_control) %>%
  group_by(g_i) %>%
  summarise(PO_0 = mean(PO_0)) %>% pander
```



$$\begin{aligned}\bar{CE}^I(\phi, \psi)&=\bar{Y}(0;\phi)-\bar{Y}(0;\psi)\\
    &=\sum_{i=3}^4\frac{\bar{Y}_i(0;\phi)}{2}-\sum_{i=1}^2\frac{\bar{Y}_i(0;\psi)}{2}\\&=0.116-0.13\\&=-0.0144\end{aligned}$$



### Total Effect

```{r}
summary %>% mutate(PO_1 = Cases_treated/Total_treated, 
                   PO_0 = Cases_control/Total_control) %>%
  group_by(g_i) %>%
  summarise(PO_1 = mean(PO_1), PO_0 = mean(PO_0)) %>% pander
```



$$\begin{aligned}\bar{CE}^T(\phi,\psi)&=\bar{Y}(0;\phi)-\bar{Y}(1;\psi)\\&=\sum_{i=3}^4\frac{\bar{Y}_i(0;\phi)}{3}-\sum_{i=1}^2\frac{\bar{Y}_i(1;\psi)}{2}\\&=0.116-0.879=-0.763\end{aligned}$$


$$\begin{aligned}\bar{CE}^T(\phi,\psi)&=\bar{CE}^I(\phi,\psi)+\bar{CE}^D(\psi)\\&=-0.0144-0.749\\&=-0.763\end{aligned}$$


### Overall Effect

```{r}
summary %>%
  mutate(PO = (Cases_treated + Cases_control)/(Total_treated + Total_control)) %>%
  group_by(g_i) %>%
  summarise(PO = mean(PO)) %>% pander
```

$$\begin{aligned}\bar{CE}^O(\phi,\psi)&=\bar{Y}(\phi)-\bar{Y}(\psi)\\&=\sum_{i=3}^4\frac{\bar{Y}_i(\phi)}{2}-\sum_{i=1}^2\frac{\bar{Y}_i(\psi)}{2}\\&=0.116-0.505=-0.389\end{aligned}$$


# 3. Variance and Confidence Interval

### Direct Effect


```{r}
var_WI <- data %>%
  group_by(group_idx, z) %>%
  summarise(var = var(y))

var_i1 <- var_WI %>% filter(z == 1) %>% .$var
var_i0 <- var_WI %>% filter(z == 0) %>% .$var

var_CE_D_given_S <- var_i1/K + var_i0/(1000-K)

var_D_psi <- data_CE_D %>% 
  filter(g == 1) %>%
  .$CE_D %>% var

var_CE_D_psi <- (1-C/N)*var_D_psi/C + sum(var_CE_D_given_S*g)/(C*N)
var_CE_D_psi

var_D_phi <- data_CE_D %>%
  filter(g == 0) %>%
  .$CE_D %>% var

var_CE_D_phi <- (1-(N-C)/N)*var_D_phi/(N-C) + sum(var_CE_D_given_S*(1-g))/(C*(N-C))
var_CE_D_phi
```

$$\begin{aligned}\hat{Var}\bigg(\hat{CE}^D(\psi)\bigg)&=(1-\frac{C}{N})\frac{\hat{\sigma}_D^2(\psi)}{C}+\frac{1}{CN}\sum_{i=1}^N\hat{Var}\bigg(\hat{CE}_i^D(\psi)|S_i=1\bigg)S_i\\&=0.0001343808\end{aligned}$$

$$\hat{\sigma}_D^2(\psi)=\sum_{i=1}^N\bigg(\hat{CE}_i^D(\psi)-\hat{CE}^D(\psi)\bigg)^2S_i\bigg/(C-1)$$

$$\hat{Var}\bigg(\hat{CE}_i^D(\psi)|S_i=1\bigg)=\frac{\hat{\sigma}_{i1}^2(\psi)}{K_i}+\frac{\sigma_{i0}^2(\psi)}{n_i-K_i}$$


$$\hat{\sigma}_{i1}^2(\psi)=\sum_{j=1}^{n_i}\{Y_{ij}(1;\psi)-\hat{Y}_{i}(1;\psi)\}^2Z_{ij}/(K_i-1)$$

$$\hat{\sigma}_{i0}^2(\psi)=\sum_{j=1}^{n_i}\{Y_{ij}(0;\psi)-\hat{Y}_{i}(0;\psi)\}^2(1-Z_{ij})/(n_i-K_i-1)$$

```{r}
CE_D_psi <- CE_D %>% filter(g == 1) %>% .$CE_D
CE_D_phi <- CE_D %>% filter(g == 0) %>% .$CE_D

c(CE_D_psi - 1.96 * sqrt(var_CE_D_psi), CE_D_psi + 1.96 * sqrt(var_CE_D_psi))
c(CE_D_phi - 1.96 * sqrt(var_CE_D_phi), CE_D_phi + 1.96 * sqrt(var_CE_D_phi))
```


# Reference

Hudgens, M. G., & Halloran, M. E. (2008). Toward causal inference with interference. Journal of the American Statistical Association, 103(482), 832-842.

Liu, L., & Hudgens, M. G. (2014). Large sample randomization inference of causal effects in the presence of interference. Journal of the American Statistical Association, 109(505), 288-301.























